---
title: "STA207 Final code"
author: "Muhan Yang"
date: "2026-03-16"
output: html_document
---

```{r ref_code_run_hide_all, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
load("/Users/mhy/Desktop/STA207/Data/df.rda")
library(tidyverse)
library(gt)
```


```{r, echo= FALSE, warning=FALSE}
#EDA based on a few observations from the Users Guide
# 1. only 26.6% participated for all four years

all4 <- df %>% filter(flaggk+flagg1+flagg2+flagg3==4)

# i get 2863, 24.67%

# for students who participate for 4 years with the same class type

all4_same_class <- df %>% 
  filter(flaggk+flagg1+flagg2+flagg3==4) %>% 
  filter(gkclasstype == g1classtype & g1classtype == g2classtype & g2classtype == g3classtype) 

# n=1556, 13.4%

# for students who participate for 3 years with the same class type

all3_sameclass <- df %>% 
  filter(flagg1+flagg2+flagg3==3) %>% 
  filter(g1classtype == g2classtype & g2classtype == g3classtype) 

all3_sameclass_same_school <- df %>% 
  filter(flagg1+flagg2+flagg3==3) %>% #in for all 3 grades
  filter(g1classtype == g2classtype & g2classtype == g3classtype) %>%
  filter(g1schid == g2schid & g2schid == g3schid) #same school each year


# n=3423, 29.5%
# this seems to be the right mix of selective and inclusive

summary_table <- tibble(
  group = c("Full Dataset",
            "K through 4th grade", 
            "K through 4th grade, same class type", 
            "1st through 3rd grade, same class type",
            "1st through 3rd grade, same class type, same school"),
  n = c(nrow(df),
        nrow(all4),
        nrow(all4_same_class), 
        nrow(all3_sameclass),
        nrow(all3_sameclass_same_school)),
  percent = c(100,
              nrow(all4)/nrow(df)*100, 
              nrow(all4_same_class)/nrow(df)*100, 
              nrow(all3_sameclass)/nrow(df)*100,
              nrow(all3_sameclass_same_school)/nrow(df)*100)
)

gt(summary_table) %>% 
  fmt_number(columns = vars(percent), decimals = 2) %>% 
  cols_label(group = "Group", n = "Number of Students", percent = "Percentage of Total") 

```


```{r ref_code_run_hide_all, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# 2. there are different scores to use, read/math/listening for SAT, also read/math BSF. 
# each of these are available for each year. 

score_cols_sat <- c('g1treadss', 'g1tmathss', 'g1tlistss', 
                'g2treadss', 'g2tmathss', 'g2tlistss', 
                'g3treadss', 'g3tmathss', 'g3tlistss')

score_cols_bsf <- c('g1readbsraw', 'g1mathbsraw',
                    'g2readbsraw', 'g2mathbsraw',
                    'g3readbsraw', 'g3mathbsraw')

sat_scores <- df %>% 
  select(all_of(score_cols_sat)) %>% 
  pivot_longer(cols = everything(), names_to = "score_type", values_to = "score") %>%
  mutate(
    class = case_when(
      str_detect(score_type, "read") ~ "reading",
      str_detect(score_type, "math") ~ "math",
      str_detect(score_type, "list") ~ "listening"),
    grade = as.factor(case_when(
      str_detect(score_type, "g1") ~ 1,
      str_detect(score_type, "g2") ~ 2,
      str_detect(score_type, "g3") ~ 3
    ))
  ) 


bsf_scores <- df %>%
  select(all_of(score_cols_bsf)) %>%
  pivot_longer(cols = everything(), names_to = "score_type", values_to = "score") %>%
  mutate(
    class = case_when(
      str_detect(score_type, "read") ~ "reading",
      str_detect(score_type, "math") ~ "math"),
    grade = as.factor(case_when(
      str_detect(score_type, "g1") ~ 1,
      str_detect(score_type, "g2") ~ 2,
      str_detect(score_type, "g3") ~ 3
    ))
  ) 


#sat_plot <- ggplot(sat_scores,
      # aes(x = class, y = score, fill=grade)) +
  #geom_violin() +
  #theme_minimal(); sat_plot

# 
#bsf_plot <-ggplot(bsf_scores,
                  #aes(x = class, y = score, fill=grade)) +
  #geom_violin() +
  #theme_minimal(); bsf_plot

#hist(sat_scores$score)
#hist(bsf_scores$score)

#par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))

#hist(sat_scores$score, main = "SAT Scores", xlab = "Score")
#hist(bsf_scores$score, main = "BSF Scores", xlab = "Score")
#observations: scores differ per class and per grade with a trend
# as grade increases, scores increase, so either control or include in design
# SAT scores are more nicely distributed than BSF scores
# choose to use SAT scores
```


```{r ref_code_run_hide_all, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
load("/Users/mhy/Desktop/STA207/Data/df_subset (2).rda")

# Convert column "subject" into factor
df_subset$subject <- as.factor(df_subset$subject)
head(df_subset)
library(dplyr)
library(car)
library(lme4)
library(influence.ME)
full_model <- lm(score ~ class_type+subject+gender+race+grade+urban, df_subset)

inter_full <- lm(score ~ class_type+subject+gender+grade+urban+race+subject*grade, df_subset)

par(mfrow = c(1, 2),
    mar = c(4, 4, 2, 1),
    cex = 0.85,
    cex.axis = 0.8,
    cex.lab = 0.9,
    cex.main = 1)
#plot(full_model, which=1)
inter_full <- lm(score ~ class_type+subject+gender+grade+urban+race+subject*grade, df_subset)

#Anova(inter_full, type="III")

mixed_model_sch <- lmer(score
    ~ class_type
    + subject
    + gender
    + race
    + grade
    + subject*grade
    + (1|schid),
    data=df_subset)
library(knitr)

random_table <- data.frame(
  Groups = c("stdntid", "Residual"),
  Name = c("(Intercept)", ""),
  Variance = c(139.8, 1363.9),
  Std.Dev. = c(11.82, 36.93)
)
library(knitr)

random_table <- data.frame(
  Groups = c("stdntid", "Residual"),
  Name = c("(Intercept)", ""),
  Variance = c(857.9, 633.9),
  Std.Dev. = c(29.29, 25.18)
)

#plot(final_model, type = c("p", "smooth"))

res <- residuals(final_model)

#Plot the histogram
hist(res, 
     breaks = 50, 
     col = "skyblue", 
     xlim = c(-100, 100),
     main = "Histogram of Residuals", 
     xlab = "Residuals")

x <- seq(min(res), max(res), length = 100)
y <- dnorm(x, mean = mean(res), sd = sd(res))/3
y <- y * diff(hist(res, plot = FALSE)$mids[1:2]) * length(res)
lines(x, y, col = "red", lwd = 2)



# Plot the histogram for random effect: student
random_effects <- ranef(final_model)
student_intercepts <- random_effects$stdntid[, "(Intercept)"]
hist(student_intercepts, 
     breaks = 20, 
     col = "darkseagreen", 
     main = "Histogram of Random Intercepts (Students)", 
     xlab = "Random Effect Estimate")



library(broom.mixed)
tidy(final_model)
categories <- list(
  subject = c("math", "reading", "listening"),
  race = c("white", "black"),
  urban = c("urban", "suburban", "inner city"),
  gender = c("male", "female"),
  grade = c("1", "2", "3")
)


library(gt)

model_table <- data.frame(
  Model = c(
    "Multiple Linear Regression\nModel without interaction term",
    "Multiple Linear Regression\nModel with interaction term",
    "Mixed Effect Model\n(school as random effect)",
    "Mixed Effect Model\n(student as random effect)"
  ),
  `Variance of Residual` = c(1517, 1472, 1364, 634),
  `Coefficient of small class` = c(8.867, 8.879, 9.468, 8.915),
  check.names = FALSE
)

gt(model_table) |>
  cols_label(
    Model = "",
    `Variance of Residual` = "Variance of Residual",
    `Coefficient of small class` = "Coefficient of small class"
  ) |>
  fmt_number(
    columns = `Coefficient of small class`,
    decimals = 3
  ) |>
  tab_style(
    style = list(
      #cell_fill(color = "#37474F"),
     # cell_text(color = "white", size = px(18))
    ),
    locations = cells_body()
  ) |>
  tab_style(
    style = list(
     # cell_fill(color = "#37474F"),
      #cell_text(color = "white", weight = "bold", size = px(20))
    ),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.border.top.color = "white",
    table.border.bottom.color = "white",
    column_labels.border.top.color = "white",
    column_labels.border.bottom.color = "white",
    table_body.hlines.color = "white",
    table_body.vlines.color = "white",
    column_labels.vlines.color = "white",
    heading.border.bottom.color = "white"
  )


# Get a list of all unique school IDs
all_schools <- unique(df_subset$schid)
# Create a vector to store results
results <- numeric(length(all_schools))

for(i in 1:length(all_schools)) {
  # Excluding one school
  temp_data <- subset(df_subset, schid != all_schools[i])
  
  temp_model <- lmer(score ~ class_type + grade + gender + subject + race + subject*grade + (1|stdntid), data = temp_data)
  
  # Coefficient for small class size
  results[i] <- fixef(temp_model)["class_typesmall"]
}

plot(results, main="Sensitivity: Leave-One-School-Out", ylab="Small Class Coefficient")
abline(h = mean(results), col="red")


# Function to run model and extract the "Small Class" coefficient
get_effect <- function(col_name, group_val, my_form) {
  sub_data <- df_subset %>% filter(!!sym(col_name) == group_val)
  
  model <- lmer(as.formula(my_form), data = sub_data)
  
  tidy(model, conf.int = TRUE, conf.method = "Wald") %>%
    filter(term == "class_typesmall") %>%
    select(estimate, conf.low, conf.high) %>%
    mutate(subgroup = paste(col_name, ":", group_val), category = col_name)
}

subject1 <- get_effect("subject", "math", "score ~ gender + race + class_type + grade + urban + (1 | stdntid)")
subject2 <- get_effect("subject", "reading", "score ~ gender + race + class_type + grade + urban + (1 | stdntid)")
subject3 <- get_effect("subject", "listening", "score ~ gender + race + class_type + grade + urban + (1 | stdntid)")

grade1 <- get_effect("grade", "1", "score ~ gender + race + class_type + subject + urban + (1 | stdntid)")
grade2 <- get_effect("grade", "2", "score ~ gender + race + class_type + subject + urban + (1 | stdntid)")
grade3 <- get_effect("grade", "3", "score ~ gender + race + class_type + subject + urban + (1 | stdntid)")

gender1 <- get_effect("gender", "male", "score ~ grade + race + class_type + subject + urban + subject*grade + (1 | stdntid)")
gender2 <- get_effect("gender", "female", "score ~ grade + race + class_type + subject + urban + subject*grade + (1 | stdntid)")

#race1 <- get_effect("race", "asian", "score ~ grade + gender + class_type + subject + urban + subject*grade + (1 | stdntid)")
race2 <- get_effect("race", "black", "score ~ grade + gender + class_type + subject + urban + subject*grade + (1 | stdntid)")
#race3 <- get_effect("race", "hispanic", "score ~ grade + gender + class_type + subject + urban + subject*grade + (1 | stdntid)")
#race4 <- get_effect("race", "other", "score ~ grade + gender + class_type + subject + urban + subject*grade + (1 | stdntid)")
race5 <- get_effect("race", "white", "score ~ grade + gender + class_type + subject + urban + subject*grade + (1 | stdntid)")

urban1 <- get_effect("urban", "inner city", "score ~ grade + gender + class_type + subject + race + subject*grade + (1 | stdntid)")
urban2 <- get_effect("urban", "rural", "score ~ grade + gender + class_type + subject + race + subject*grade + (1 | stdntid)")
urban3 <- get_effect("urban", "suburban", "score ~ grade + gender + class_type + subject + race + subject*grade + (1 | stdntid)")
urban4 <- get_effect("urban", "urban", "score ~ grade + gender + class_type + subject + race + subject*grade + (1 | stdntid)")

all_results <- bind_rows(subject1, subject2, subject3,
    grade1, grade2, grade3,
    gender1,